---
title: "Introduction to gaplightr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to gaplightr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

gaplightr estimates forest canopy light transmission by combining LiDAR point
clouds, terrain data, and hemispherical photography. Given a set of sampling
points, it generates synthetic fisheye photographs and computes solar radiation
metrics such as canopy openness, transmitted global irradiation, and the light
penetration index.

The package is designed for batch workflows on large stream networks. Each
pipeline step writes output files to disk because the intermediate files are useful on
their own and so interrupted runs can resume from where they left off using `resume = TRUE`.

## Inputs

We will use a small internal demo dataset to demonstrate a basic gaplightr pipeline:

- a 1 km x 1 km DEM,
- three sampling points
- a matching LiDAR tile

```{r load-demo-data}
library(gaplightr)
library(terra)
library(sf)

dem_path <- system.file("extdata", "dem.tif", package = "gaplightr")
points_path <- system.file("extdata", "points.geojson", package = "gaplightr")
lidar_dir <- system.file("extdata", "lidar", package = "gaplightr")

output_dir <- tempdir()
```

```{r plot-inputs}
#| fig.alt: "DEM with three sampling points overlaid"
#| fig.width: 7
#| fig.height: 5
dem <- terra::rast(dem_path)
pts <- sf::read_sf(points_path)

terra::plot(dem, main = "Demo DEM with sampling points")
plot(pts, add = TRUE, pch = 21, bg = "white", cex = 1.5)
```

## Step 1: Load points

`gla_load_points()` validates the CRS, extracts elevation from the DEM, and
assigns a `point_id` used to name all downstream output files.

```{r load-points}
points <- gla_load_points(points_path, dem_path)
str(points)
```

## Step 2: Create virtual plots

`gla_create_virtual_plots()` clips a circular LiDAR plot around each point and
saves one `.las` file per point.

```{r virtual-plots}
vp_dir <- file.path(output_dir, "vp")

points <- gla_create_virtual_plots(
  points = points,
  folder = lidar_dir,
  output_dir = vp_dir,
  plot_radius = 25,
  resume = FALSE
)

list.files(vp_dir)
```

## Step 3: Extract horizons

`gla_extract_horizons()` calculates the terrain horizon angle at each azimuth
direction and caches the result as a `_horizon.csv` file. This mask prevents
terrain from being counted as sky in the fisheye photo.

```{r extract-horizons}
hz_dir <- file.path(output_dir, "hz")

points <- gla_extract_horizons(
  points = points,
  dem_path = dem_path,
  output_dir = hz_dir,
  parallel = FALSE,
  resume = FALSE
)

list.files(hz_dir)
```

The `horizon_mask` list-column stores polar-projected horizon coordinates for
each point:

```{r inspect-horizon}
str(points$horizon_mask[[1]])
```

## Step 4: Create fisheye photos

`gla_create_fisheye_photos()` projects the clipped LiDAR onto a hemispherical
plane, overlays the horizon mask, and writes a `.bmp` image per point.

```{r create-photos}
photo_dir <- file.path(output_dir, "photos")

points <- gla_create_fisheye_photos(
  points = points,
  output_dir = photo_dir,
  camera_height_m = 1.37,
  min_dist = 1,
  max_dist = 70,
  img_res = 2800,
  max_cex = 0.5,
  min_cex = 0.03,
  pointsize = 10,
  dpi = 1200,
  parallel = FALSE,
  resume = FALSE
)

list.files(photo_dir)
```

```{r show-photo}
#| fig.alt: "Synthetic fisheye photograph showing canopy gap pattern"
#| fig.width: 7
#| fig.height: 5
photo_file <- points$fisheye_photo_path[[1]]
img <- imager::load.image(photo_file)
plot(img, axes = FALSE, main = "Synthetic fisheye photo - point 1")
```

## Step 5: Process for solar radiation

`gla_process_fisheye_photos()` reads each fisheye photo, computes gap fractions
by sky ring and sector, and integrates solar irradiance over the specified day
range. For this vignette, we use a short illustrative window (Julian days 172–182,
i.e., around the summer solstice) and a coarser time step to keep computation
fast during automated checks.

```{r process-photos}
results <- gla_process_fisheye_photos(
  points = points,
  clearsky_coef = 0.65,
  time_step_min = 10,
  day_start = 172,
  day_end = 182,
  day_res = 2,
  elev_res = 5,
  azi_res = 5,
  Kt = 0.54,
  parallel = FALSE
)

str(results)
```

Setting `keep_gap_fraction_data = TRUE` retains the full gap fraction matrix for
each point as a list-column, which is useful for diagnostic plots or custom
downstream analyses.

```{r process-photos-gap-fraction}
results_with_gf <- gla_process_fisheye_photos(
  points = points,
  clearsky_coef = 0.65,
  time_step_min = 10,
  day_start = 172,
  day_end = 182,
  day_res = 2,
  elev_res = 5,
  azi_res = 5,
  Kt = 0.54,
  keep_gap_fraction_data = TRUE,
  parallel = FALSE
)

str(results_with_gf$gap_fraction_data[[1]])
```

## Next steps

For larger datasets, enable parallel processing by configuring a `future` plan
before any pipeline step:

```r
future::plan(future::multisession, workers = 4)

points <- gla_extract_horizons(
  points     = points,
  dem_path   = dem_path,
  output_dir = hz_dir,
  parallel   = TRUE,
  resume     = TRUE   # skip points already processed
)
```

The `resume = TRUE` default means you can safely interrupt and restart any batch
step without reprocessing completed points.
