---
title: "Analysing a Real Fisheye Photo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysing a Real Fisheye Photo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette demonstrates the real-photo workflow: loading a raw hemispherical
photograph, preprocessing it for gap fraction analysis, and computing solar
irradiance metrics. This path is an alternative to the synthetic LiDAR-based
fisheye pipeline shown in the introduction vignette.

The bundled photo `example-photo.jpg` was taken with a Sigma 8mm fisheye lens
mounted on a camera body held at breast height under a forest canopy.

## Step 1: Load the raw photo

```{r load-photo}
#| fig.alt: "Raw hemispherical photograph before preprocessing"
#| fig.width: 7
#| fig.height: 5
library(gaplightr)
library(imager, warn.conflicts = FALSE)

img_path <- system.file("extdata", "example-photo.jpg", package = "gaplightr")
img <- imager::load.image(img_path)
plot(img, axes = FALSE, main = "Raw fisheye photo")
```

## Step 2: Preprocess the photo

Real fisheye photos require several preprocessing steps before gap fraction can
be extracted. The exact parameters are photo-specific and must be determined
visually for each image:

- **Crop**: remove the camera body and any non-circular border from the frame
- **Rotate**: rotate counterclockwise to align true north to the top
- **Channel extraction**: isolate the blue channel, which gives the best
  contrast between sky and canopy
- **Threshold**: convert to a binary sky/canopy mask

```{r preprocess}
#| fig.alt: "Binary sky/canopy mask after thresholding"
#| fig.width: 5
#| fig.height: 5

# Crop to the circular fisheye area. These pixel bounds are specific to
# example-photo.jpg (2184 x 1456 px) and were determined by visual inspection.
img_crop <- imager::imsub(
  img,
  x %inr% c(372, 1818),
  y %inr% c(2, 1448)
)

# Rotate fisheye 90 degrees so that true north is up. Nearest-neighbour
# interpolation avoids mixing sky and canopy pixel values at boundaries.
img_rotated <- imager::imrotate(img_crop, 90, interpolation = 0)

# Split RGB and keep the blue channel (best contrast)
img_blue <- imager::channel(img_rotated, 3)

# Threshold to produce a binary sky mask. The adjust value of 1.75 scales the
# auto-detected threshold and is specific to this photo; images taken under
# different conditions will need re-tuning.
img_thresh <- imager::threshold(img_blue, thr = "auto", adjust = 1.75)
fisheye <- imager::as.cimg(img_thresh)

plot(fisheye, axes = FALSE, main = "Thresholded fisheye (white = sky)")
```

Save the preprocessed image as a BMP file. BMP is required by
`gla_process_fisheye_photos()`.

```{r save-bmp}
preprocessed_path <- file.path(tempdir(), "example-photo-preprocessed.bmp")
w <- imager::width(fisheye)
h <- imager::height(fisheye)
grDevices::bmp(
  preprocessed_path,
  width = w,
  height = h,
  units = "px",
  res = 72,
  type = "cairo"
)
graphics::par(mar = c(0, 0, 0, 0))
plot(fisheye, axes = FALSE)
grDevices::dev.off()
```

## Step 3: Load a point location

`gla_load_points()` reads a spatial points file, reprojects to match the DEM,
and extracts elevation and coordinates. We take a single point and attach the
path to the preprocessed fisheye image.

```{r load-point}
dem_path <- system.file("extdata", "dem.tif", package = "gaplightr")
points_path <- system.file("extdata", "points.geojson", package = "gaplightr")

points <- gla_load_points(points_path, dem_path)
points <- points[1, ]
points$fisheye_photo_path <- preprocessed_path

str(points)
```

## Step 4: Inspect gap fractions

Before computing irradiance, it is useful to inspect the gap fraction grid
directly. `gla_extract_gap_fraction()` maps sky-visible pixels into rings
(elevation bands) and sectors (azimuth wedges).

The Sigma 8mm lens uses an equisolid-angle projection rather than the
equidistant polar projection assumed by default. `gla_lens_sigma_8mm()`
returns the corresponding radial distortion calibration.

`rotation_deg = 16` is the clockwise angular offset between camera orientation
and true north for this particular photo.

```{r gap-fraction}
sigma_cal <- gla_lens_sigma_8mm()

gap_result <- gla_extract_gap_fraction(
  img_file = preprocessed_path,
  elev_res = 5,
  azi_res = 5,
  rotation_deg = 16,
  radial_distortion = sigma_cal
)

gap_result
```

## Step 5: Compute solar irradiance

`gla_process_fisheye_photos()` integrates gap fractions with the solar geometry
model to produce canopy openness and irradiance metrics. We use Julian days
172-182 (around the summer solstice) with a coarser time step to keep
computation fast.

```{r process-photos}
results <- gla_process_fisheye_photos(
  points = points,
  clearsky_coef = 0.65,
  time_step_min = 10,
  day_start = 172,
  day_end = 182,
  day_res = 2,
  elev_res = 5,
  azi_res = 5,
  Kt = 0.54,
  rotation_deg = 16,
  radial_distortion = sigma_cal,
  parallel = FALSE
)

str(results[, c(
  "canopy_openness_pct",
  "transmitted_global_irradiation_MJm2d",
  "light_penetration_index"
)])
```

## Next steps

For a batch of real photos, create an sf object with one row per photo and set
`fisheye_photo_path` to the preprocessed BMP path for each row. The same
`gla_process_fisheye_photos()` call then processes all points.